Tests of measurement invariance without subgroups: A~generalization of classical methods

### 2012-02-11 www/rev1/response.txt

Comment 1.2. There are clear links between the current approach based on an
observed continuous grouping variable V and the factor mixture approach. In
factor mixture modeling, the grouping variable that characterizes the subgroups
that differ in terms of particular model parameters are also typically unknown
(i.e., latent classes), but class membership can be predicted on the basis of
observed predictor(s). So the model in Figure 1 and the scenario of 2 subgroups
that the authors describe on pages 17-20 is essentially a factor mixture model
with two latent classes in which age predicts class membership. Therefore, the
authors should take steps to relate their modeling approach to factor mixture
modeling.

Response 1.2. We agree and have now described our methods' relationship to
mixture models in the Discussion.  Some advantages of our approach include the
facts that it does not require a priori specification of the number of latent
classes or a priori specification of the way by which age impacts class
membership.

Comment 1.3. Similarly, there exist a family of moderation-type factor models
that may serve the same purpose of studying failures of invariance in common
factor models with respect to (unknown) "grouping" variables. For instance,
Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models
that were based on Shaun Purcell's work in which certain parameters are
moderated by either observed or unobserved variables. These models too could be
applied to the important problems related to invariance with respect to unknown
groups although they may not work as well as do mixture models and the current
approach when indeed there is some step function that describes the violations
of invariance. The existence of these alternative approaches by no means
diminishes the potential value of the current developments, but it does mean
that the authors should discuss the relation between their approach and other
existing methods.

Response 1.3. We now describe factor analysis models employing moderation in the
Discussion.

@@

Response 3.2. We have now incorporated these results in a new section, titled
"Local alternatives".  We also discuss these results in the context of the
simulation, showing the impact of our imposed measurement invariance violation
on the Brownian bridge.

We addressed all minor issues identified by Reviewer 3.  We have a response to
one of these issues:

Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly
specified is bonus when the second point above is incorporated. I think this
would improve the manuscript.

Response 3.3. Violations of distributional assumptions may still cause problems
with the proposed tests, even if we need not worry about the "structure" of the model (e.g., parameter constraints, included variables, etc).  We have modified this section of text accordingly.

@@

Comment 4.4. Page 19:	The overall k=19 model parameter consists of six factor
loadings, six measurement errors and one covariance between the factors. The
authors refer to the six means of the manifest variables. Usually, in factor
models mean differences of the factors are of research interest. The authors
should give some information why they considered the manifest means instead of
the latent mean differences.

Response 4.4. Factor means can be estimated if one is constraining other model
parameters to be equal across groups.  In our new data example, we do estimate
factor means and use them in the proposed tests.  In our simulation we use a
scenario where no grouping variables exist (i.e., there is only a continuous
V).  In this scenario, it seems difficult to estimate factor means prior to model estimation because we cannot constrain parameters across groups.  We have added some extra detail to the simulation section on our use of a single-group model.

Comment 4.5.	The simulations later on focus on invariance of factor loadings.
Basicly, the covariance structure in the simulations are analyzed with the
factor model of Figure 1. So, means are not included in the analyses (as far as
I read the simulation results correctly).

Response 4.5. Mean parameters are included in the casewise likelihood function,
though they were unmodeled in the simulation and thus omitted from Figure 1. 
The proposed tests are most easily carried out using the casewise likelihood function.  Also see Comment/Response 1.5.

### 2012-02-11 www/paper.Rnw
changed paper title to better reflect that there are no predefined subgroups and that the tests extend classical approaches
updated acknowledgements (added Yves and Psychoco)

further small improvements



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@105 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Generalized measurement invariance tests with application to
  factor analysis}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

\abstract{
  The issue of measurement invariance commonly arises in factor-analytic
  contexts, with methods for assessment including likelihood ratio tests,
  Lagrange multiplier tests, and Wald tests. These tests all require advance
  definition of the number of groups, group membership, and offending model
  parameters. In this paper, we construct tests of measurement invariance based on
  stochastic processes of casewise derivatives of the likelihood function. These
  tests can be viewed as generalizations of the Lagrange multiplier test, and they
  are especially useful for: (1)~identifying specific parameters impacted by
  measurement invariance violations, and (2)~identifying subgroups of individuals
  that violated measurement invariance based on a continuous auxiliary variable. 
  The tests are presented and illustrated in detail, including an application
  to a study of stereotype threat and simulations examining the tests'
  abilities in controlled conditions.
}

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat application.
  Correspondence to Edgar C.\ Merkle, 
  Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Generalized measurement invariance tests}
\rightheader{Generalized measurement invariance tests}

@@

To carry out
the tests, we first fit the four-group model to the data,
calculating casewise derivatives and 
the observed information matrix.  We then
calculated the cumulative score process with respect to student GPA via
\eqref{eq:cumscore}, allowing us to
obtain various test statistics and $p$-values from this 
process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
statistic from \eqref{eq:maxlm}.

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change: we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
The students are ordered according to GPA (i.e.,  in this application the variable $V$),
and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.

>>>>

\title{Tests of measurement invariance without subgroups:
  A~generalization of classical methods}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

\abstract{
  The issue of measurement invariance commonly arises in factor-analytic
  contexts, with methods for assessment including likelihood ratio tests,
  Lagrange multiplier tests, and Wald tests. These tests all require advance
  definition of the number of groups, group membership, and offending model
  parameters. In this paper, we construct tests of measurement invariance based on
  stochastic processes of casewise derivatives of the likelihood function. These
  tests can be viewed as generalizations of the Lagrange multiplier test, and they
  are especially useful for: (1)~identifying subgroups of individuals
  that violate measurement invariance along a continuous auxiliary variable
  without prespecified thresholds, and (2)~identifying specific parameters impacted by
  measurement invariance violations.
  The tests are presented and illustrated in detail, including an application
  to a study of stereotype threat and simulations examining the tests'
  abilities in controlled conditions.
}

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat application,
  Yves Rosseel, who provided feedback and code for performing the
  tests with the lavaan package, and the participants of the
  Psychoco 2012 workshop on psychometric computing for
  discussions.
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Tests of measurement invariance without subgroups}
\rightheader{Tests of measurement invariance without subgroups}

@@

To carry out the tests, we first fit the four-group model to the data,
calculating casewise derivatives and  the observed information matrix.   Second,
to assess measurement invariance within the ``minority, stereotype prime'' group
only, the scores of the $n_{\text{MSP}}$ individuals from that group are ordered
and aggregated along GPA (i.e.,  in this application the variable $V$). The
cumulative score process \eqref{eq:cumscore} with respect to GPA then allows us
to obtain various test statistics and $p$-values from this  process.  Test
statistics include the double-max statistic from \eqref{eq:dmax}, the
Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
statistic from \eqref{eq:maxlm}.

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change: we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.
